---
title: "Spec-first workflow with vision"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spec-first workflow with vision}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(vision)
```

## Introduction

`vision` enables spec-first workflows for consumer psychology research. Instead of writing analysis code first, you define your complete study specification in a YAML or JSON file. `vision` then validates this specification and constructs an auditable analysis recipe.

This approach ensures:

- **Reproducibility**: All analysis decisions are recorded before touching data
- **Auditability**: Complete audit trail from specification to results
- **Peer-review readiness**: Specifications serve as transparent documentation

## Schema v0.1.0

`vision` validates specifications against schema version 0.1.0. This schema requires seven top-level sections:

### Required Sections

1. **meta**: Study metadata
   - `title`: Study title (character)
   - `authors`: Author list (character vector or list)
   - `created`: Creation date (character or Date)

2. **data**: Data source configuration
   - `sources`: File paths (character vector)
   - `id`: ID column name (character)
   - `joins`: Join specifications (optional list)

3. **vars**: Variable specifications
   - `rename`: Variable renaming map (optional named character vector)
   - `types`: Variable types (named character vector with values: "character", "double", "integer", "logical", "factor", "date", "datetime")

4. **rules**: Data processing rules
   - `exclusions`: Exclusion criteria (optional list)
   - `missingness`: Missingness handling (optional list)

5. **scales**: Scale definitions
   - Each scale has: `name`, `items`, `reverse` (optional), `min_items`, `composite`

6. **models**: Statistical model specifications
   - Each model has: `name`, `formula`, `family` (optional), `robust` (optional)

7. **outputs**: Output configuration
   - `root`: Output directory path (character)
   - `render`: Rendering specifications (optional list)

## Workflow

### 1. Create a template

Start by generating a minimal template:

```{r template}
template_path <- tempfile(fileext = ".yml")
write_spec_template(template_path)
```

The template includes all required sections with commented examples:

```{r show_template, comment=''}
cat(readLines(template_path), sep = "\n")
```

### 2. Read the specification

Parse the YAML file into a `niche_spec` object:

```{r read}
spec <- read_spec(template_path)
class(spec)
```

If the specification file doesn't include a `schema_version` field, `read_spec()` automatically sets it to the current version.

### 3. Validate the specification

Validate against schema v0.1.0 requirements:

```{r validate}
spec <- validate_spec(spec)
```

Validation checks:

- All required top-level fields exist
- All required subfields exist and have correct types
- Variable types use allowed values
- Scales and models have required fields

Validation is **structural**, not semantic—it ensures the specification is well-formed but doesn't verify that variable names exist in data or that formulas are valid R syntax.

### 4. Build a recipe

Construct an auditable recipe from the validated specification:

```{r build}
recipe <- build_recipe(spec)
class(recipe)
```

The recipe includes:

- **spec_hash**: Deterministic hash of the canonicalized specification
- **defaults_applied**: Tracking of all applied default values
- **outputs**: Created output directory structure
- **created**: ISO-8601 timestamp

### 5. Write the recipe

Serialize the recipe to canonical JSON:

```{r write, eval=FALSE}
recipe_path <- write_recipe(recipe)
recipe_path
```

By default, the recipe is written to `outputs/audit/recipe.json`. You can specify a custom path:

```{r write_custom, eval=FALSE}
custom_path <- file.path(tempdir(), "my_recipe.json")
write_recipe(recipe, path = custom_path)
```

## Deterministic Hashing

`build_recipe()` canonicalizes specifications before hashing to ensure determinism:

```{r hash_demo}
# Create two equivalent specs with different ordering
spec1 <- spec
spec2 <- spec

# Different variable ordering
spec1$vars$types <- c(participant_id = "character", age = "integer")
spec2$vars$types <- c(age = "integer", participant_id = "character")

# Different author format
spec1$meta$authors <- c("Author 1", "Author 2")
spec2$meta$authors <- list("Author 1", "Author 2")

# Build recipes
recipe1 <- build_recipe(spec1)
recipe2 <- build_recipe(spec2)

# Hashes are identical
identical(recipe1$spec_hash, recipe2$spec_hash)
```

Canonicalization includes:

- Converting `meta$authors` lists to character vectors
- Sorting `vars$rename` and `vars$types` by name
- Sorting `scales` and `models` by name
- Ensuring `reverse` fields exist (as `character(0)` if missing)

## Default Tracking

`build_recipe()` tracks all applied defaults:

```{r defaults_demo}
# Create minimal spec (many optional fields missing)
minimal_spec <- spec
minimal_spec$data$joins <- NULL
minimal_spec$vars$rename <- NULL
minimal_spec$rules$exclusions <- NULL
minimal_spec$rules$missingness <- NULL
minimal_spec$outputs$render <- NULL

recipe_minimal <- build_recipe(minimal_spec)
names(recipe_minimal$defaults_applied)
```

Defaults are applied ONLY to optional fields and recorded with their exact field paths.

## Error Messages

`vision` provides actionable error messages for specification problems:

```{r error_demo, error=TRUE}
# Missing required field
bad_spec <- spec
bad_spec$meta$title <- NULL

validate_spec(bad_spec)
```

```{r error_demo2, error=TRUE}
# Invalid variable type
bad_spec2 <- spec
bad_spec2$vars$types <- c(participant_id = "invalid_type")

validate_spec(bad_spec2)
```

All errors include:

- What failed
- Where it failed (exact field path)
- How to fix it

## Next Steps

`vision` prepares auditable recipes but does not execute analyses. For complete workflows:

- Use `nicheData` (future) for data ingestion
- Use `nicheScales` (future) for scale scoring
- Use `nicheModels` (future) for statistical modeling
- Use `nicheReport` (future) for APA-formatted outputs

The `niche` meta-package will provide a unified interface to the complete workflow.

## Cleanup

```{r cleanup, include=FALSE, eval=FALSE}
# Clean up temporary files
# Note: Some paths may not exist if chunks were not evaluated
if (exists("template_path")) unlink(template_path)
if (exists("custom_path")) unlink(custom_path)
if (exists("recipe_path")) unlink(dirname(recipe_path), recursive = TRUE)
```
